Safe Memory Regions for Big Data Processing

### 2016-06-07 paper/type-inference.tex
type inference

\subsection{Region Parametrization/Instantiation for Classes}
\label{sec:fb-templatization}

The first phase is an iterative process involving the following three steps,
the first two of which are mutually dependent on each other.

\emph{Region Parametrization}. For every class \C{C}, we identify
a sequence of formal region parameters $\pi_0, \cdots \pi_n$ that \C{C}
should be parametric over.

\emph{Parameter Instantiation}.
We then replace every instance of class \C{C} in the program by an instance
$\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots, \rho_n$
are fresh identifiers denoting actual region parameters.

@@

\begin{tabular}{ccc}
\begin{minipage}{0.28\linewidth}
\begin{codejava}
class List
  $\extends$ $\ObjZ$
{
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow$
&
\begin{minipage}{0.5\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$
  $\extends$ $\ObjZ \langle \rho_0 \rangle$
{
  $\ObjZ \langle \rho_1 \rangle$ data;
  List $\langle \rho_0, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}

@@

\emph{Type-Parametric Classes}.
The techniques described above can be used for parametrized classes
with the following extension. Consider a parametrized class \C{C $\langle$ T $\extends$ B $\rangle$}.
The parametric type \C{T} is instantiated using the number of region parameters
that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
to be \C{Object}, and \C{T} is instantiated with one region parameter.

@@

\subsection{Region Parametrization/Instantiation for Methods}

@@

Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
%% over a set of fixed formal region parameters,
where the set of region-constraints $\phi$ is defined by:
\begin{smathpar}
\begin{array}{lcl}
\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
\end{array}
\end{smathpar}

Our constraints also make uses of \emph{pending substitutions}\footnote{we borrowed this terminology from~\cite{ltpldi08}} $F$.
A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
used in a particular context:
\begin{smathpar}
\begin{array}{lcl}
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
For instance, in the validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.

@@

\item Validity constraints of form $\isvalid{\phictxt \conj \varphi_i} {\phicstr}$
where $\varphi_i$ is a predicate variable (representing the precondition of a
method to be determined), $\phicstr$ is a region constraint that is \emph{required}
to hold at a particular program point (within the method), and $\phictxt$ is
a region constraint that is \emph{known} to hold at that program point.
Each $\phi$ is a region constraint over region variables and region parameters:
\begin{smathpar}
\begin{array}{lcl}
\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
\end{array}
\end{smathpar}
%
%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
%% of predicate variables and pending substitutions. While $\phictxt$
%% captures relationships that are \emph{known} to hold between concrete
%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
%% $\phicstr$ captures relationships that are \emph{required} to hold
%% among region varibles (i.e., $\rho$'s), or relationships between
%% region variables and identifiers.
%

\item Validity constraints of the form $\isvalid{\phictxt \conj \varphi_i} {F_j(\varphi_j)}$
generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
$\varphi_i$ are as above).

@@

% A pending substitution ($F$) is a substitution function over region
% variables/identifiers. They represent the substitutions that need to
% be carried out when a predicate variable ($\varphi$) is replaced by a
% concrete formula in a validity constraint.

% Validity constraints primarily result from checking if a region type
% template of form $B\inang{\tbar}\inang{\rhoalloc,\rhobar}$ is
% well-formed (i.e., $\typeOk$). The corresponding well-formedness rule
% (Fig.~\ref{fig:fb-staticsem}) requires that the instantiation of $B$'s
% formal region parameters with actual region arguments satisfy its
% precondition, where actuals are subsituted for formals. Since $B$'s
% precondition could be unknown, the requirement results in a in a
% validity constraint with predicate variables and pending
% substitutions. Other sources of validity constraints are calls to
% region-polymorphic methods, where similar situation arises.

%
%% If the constraint is generated while checking the well-formedness of a
%% type or elaborating an expression, then $\varphi_j$'s denote the unknown
%% preconditions of classes and methods that were used in that type or expression.
%
% Each use of a (region-polymorphic) class or a method may instantiate region
% parameters differently, resulting in a different pending substitution
% ($F_j$).

Each region variable occuring in $\phicstr$ has an associated well-formedness
constraint, which specifies its unification domain. The unification domain of a
constraint is the union of unification domains of all region variables
occuring in the constraint.

\emph{Class Constraints}. 
Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
fairly straightforward. It is the conjunction of
\begin{itemize}
\item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
\item $\varphi_{\C{B}}$
\item $[\rho_a/\rho_f]\varphi_{\C{F}}$
\end{itemize}

@@

\paragraph{Constraint Solution}
Solving the set ($C$) of constraints entails finding an
assignment for each predicate variable $\varphi$) and each region
variable ($\rho$) that occurs free in $C$, such that the solution
satisfies well-formedness constraints on $\varphi$ and $\rhobar$.
% To simplify presentation, we think of $C$ as being parameterized on
% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
% \rbrack$. We now formalize the constraint satisfaction problem, and
% its solution.

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem (CSP)})} A constraint
satisfaction problem is a tuple $\CSP$, where $C\lbrack
\varphi,\rhobar \rbrack$ is a set of validity constraints, where
$i$'th validity constraint assumes one of the following forms:
\begin{center}
\(
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i}}\qquad
    \isvalid{\phictxt^{i} \conj \varphi}
            {F_i(\varphi)}
\)
\end{center}
$\bar{\rhoenv}$ are the unification domains for $\rhobar$. We call the
union of all unification domains ($\bigcup\bar{\rhoenv}$) as the
unification domain ($\rhoenv$) of the CSP.  The solution to the
constraint satisfaction problem is a pair $(\substFn,\phisol)$, where
$\substFn$ is a map from $\rhobar$ to $\rhoenv$ such that
$\substFn(\rho_j) \in \rhoenv_j$, for every $j$, and $\phisol$ is a
constraint formula such that:
\begin{itemize}
\item $\phisol$ is well-formed under $\rhoenv_{\varphi}$ (i.e.,
$\tywf{\rhoenv_{\varphi}}{\phisol}$).
\item Every sequent in $C\lbrack (\phisol,\substFn(\rhobar)) \rbrack$
is valid.
% \begin{center}
% \(
%   \isvalid{\phictxt^{i} \conj \bigwedge_j(\rho_j = \substFn(\rho_j)) \conj \phisol}
%           {\phicstr^{i} \conj  F_i(\phisol))}
% \)
% \end{center}
\end{itemize}
\end{definition}

@@

The first step of solving the CSP $\CSP$ is to cast it as an
equivalent problem $\csp$ involving a single validity constraint. The
single constraint ($\lbrack \varphi,\rhobar \rbrack$) is:
\begin{center}
\(
  \isvalid{ \Phicx \conj \varphi}
        {\Phics \conj  \bigwedge_i F_i(\varphi)}
\)
\end{center}
Where $\Phicx = \bigwedge_i \phictxt^{i}$ and $\Phics = \bigwedge_i
\phicstr^{i}$. To see why both CSPs are equivalent, consider two
distinct constraints, $c_i\lbrack \varphi,\rhobar_i \rbrack$ and
$c_j\lbrack \varphi,\rhobar_j \rbrack$. Since $i\neq j$, $\rhobar_i
\neq \rhobar_j$. Without the loss of generality, assume that $c_i$ was
generated before $c_j$ by the constraint generation algorithm. Observe
that when our constraint generation algorithm generates a constraint,
all relationships between concrete region identifiers referred by the
consequent of constraint are already present in its antecedent (this
includes the unification domains of region variables referred by the
consequent).  Since no new relationships between existing regions are
added when a new constraints is generated, $\phictxt^j$ either
describes the same relationships that are already described by
$\phictxt^i$, or describes relationships among region identifiers that
are new, and not relevant to $c_i$. Therefore, strengthening the
context of $c_i$ with that of $c_j$ (or vice versa) neither weakens
nor strengthens $c_i$ (resp.  $c_j$).

For the constraint sets $C_1$, $C_2$, and $C_3$ in the running
example, equivalent constraints are $c_{12}$, $c_{13}$, and $c_{14}$,
respectively, shown below:
\begin{smathpar}
\begin{array}{l}
  \csid{12} \isvalid{\varphi_0}{\rho_0 \outlives \rhoalloc_0 \conj \rho_1
     \succeq \rhoalloc_0 \conj \rho_4 = \rhoalloc_0 \conj 
     \rho_2 = \rho_0 \conj \rho_3 = \rho_1}\\
  \csid{13} \isvalid{\varphi_0 \conj \varphi_1} {\rho_5 = \rho_0}\\
  \csid{14} \isvalid{\varphi_1 \conj \varphi_2 \conj \rgn_0 \outlives \rgn_1} 
    {
        \rho_7 \outlives \rho_6 \conj \rho_8 \outlives \rho_6 \conj 
        \rho_7 = \rho_9 \conj 
    }\\
  \hspace*{1.4in} \rho_{8} = \rho_9 \conj
        [\rgn_1/\rhoalloc_2][\rgn_1/\rho_6][\rgn_0/\rho_{7-9}]\varphi_2
\end{array}
\end{smathpar}

>>>>

\subsection{Region Parametrization for Classes}
\label{sec:fb-templatization}

Region parametrization is an iterative process involving the following three steps,
the first two of which are mutually dependent on each other.

\emph{Introduction of Formal Region Parameters}.
For every class \C{C}, we identify a sequence of formal region parameters
$\pi_0, \cdots \pi_n$ that \C{C} should be parametric over.

\emph{Introduction of Actual Region Parameters}.
We then replace every instance of class \C{C} in the program by an instance
$\C{C}\langle \rho_0, \cdots, \rho_n \rangle$, where $\rho_0, \cdots, \rho_n$
are fresh identifiers denoting actual region parameters.

@@

\begin{tabular}{ccc}
\begin{minipage}{0.28\linewidth}
\begin{codejava}
class List
  $\extends$ $\ObjZ$
{
  $\ObjZ$ data;
  List next;
}
\end{codejava}
\end{minipage}
&
$\Rightarrow$
&
\begin{minipage}{0.5\linewidth}
\begin{codejava}
class List $\langle \rho_0, \rho_1 \; | \; \varphi \rangle$
  $\extends$ $\ObjZ \langle \rho_0 \rangle$
{
  $\ObjZ \langle \rho_1 \rangle$ data;
  $\C{List} \langle \rho_0, \rho_1 \rangle$ next;
}
\end{codejava}
\end{minipage}
\end{tabular}

@@

\emph{Type-Parametric Classes}.
Type parameters of classes are handled as follows.
Consider a type-parametric class \C{C $\langle$ T $\extends$ B $\rangle$}.
The parametric type \C{T} is instantiated using the number of region parameters
that its bound \C{B} has. If no bound is specified for \C{T}, the bound is taken
to be \C{Object}, and \C{T} is instantiated with one region parameter.

@@

\subsection{Region Parametrization for Methods}

@@

Every predicate variable $\varphi$ denotes an unknown \emph{region-constraint},
%% over a set of fixed formal region parameters,
where the set of region-constraints $\phi$ is defined by:
\begin{smathpar}
\begin{array}{lcl}
\phi & \coloneqq & true \ALT \rho \outlives \rho \ALT \rho = \rho \ALT \phi \conj \phi \\
\end{array}
\end{smathpar}
We will use the term \emph{validity constraint} to denote an entailment constraint
of the form $\isvalid{\varphi}{\rho_1 \outlives \rho_2}$.

Our constraints also make uses of \emph{pending substitutions}\footnote{we borrowed this terminology from~\cite{ltpldi08}} $F$.
A pending substitution serves to bind formal region parameters in $\varphi$ to the actual region parameters
used in a particular context:
\begin{smathpar}
\begin{array}{lcl}
F & \coloneqq & \cdot \ALT [\rho/\rho]F \\
\end{array}
\end{smathpar}
E.g., in the validity constraint $\isvalid{\rgn_1 \outlives
\rgn_2}{[\rgn_1/\rho_1][\rgn_2/\rho_2]\varphi}$, the pending substitution
is $[\rgn_1/\rho_1][\rgn_2/\rho_2]$. Any concrete formula (call it
$\phisol$) over variables $\rho_1$ and $\rho_2$ is a solution to
$\varphi$ if and only the formula obtained by substituting $\rgn_1$
and $\rgn_2$ for $\rho_1$ and $\rho_2$ (resp.) in $\phisol$ is
deducible from $\rgn_1 \outlives \rgn_2$.

@@

\item Validity constraints of form $\isvalid{\varphi_i \conj \phictxt} {\phicstr}$
where $\varphi_i$ is a predicate variable (representing the precondition of a
method to be determined), $\phicstr$ is a region constraint that is \emph{required}
to hold at a particular program point (within the method), and $\phictxt$ is
a region constraint that is \emph{known} to hold at that program point.

%
%% Formulas $\phictxt$ and $\phicstr$ are concrete, i.e., free
%% of predicate variables and pending substitutions. While $\phictxt$
%% captures relationships that are \emph{known} to hold between concrete
%% region identifiers (i.e., $\rgn$'s) when the constraint was generated,
%% $\phicstr$ captures relationships that are \emph{required} to hold
%% among region varibles (i.e., $\rho$'s), or relationships between
%% region variables and identifiers.
%

\item Validity constraints of the form $\isvalid{\varphi_i \conj \phictxt} {F_j(\varphi_j)}$
generated by an invocation of a method with precondition $\varphi_j$ (where $\phictxt$ and
$\varphi_i$ are as above).

@@

\paragraph{Constraint Solution}
Solving the set ($C$) of constraints entails finding an
assignment for each predicate variable $(\varphi$) and each region
variable ($\rho$) that occurs free in $C$, that satisfies all
the validity constraints as well as the well-formedness constraints on
$\varphi$ and $\rhobar$.
% To simplify presentation, we think of $C$ as being parameterized on
% $\varphi$ and $\rhobar$, and write it as $C\lbrack \varphi,\rhobar
% \rbrack$. We now formalize the constraint satisfaction problem, and
% its solution.

\begin{definition}
\emph{(\textbf{Constraint Satisfaction Problem (CSP)})}
A constraint satisfaction problem is a tuple
$(\bar{\pi}, \bar{\rho}, \bar{\varphi}, \regionDeltaMap, \predDeltaMap, C)$,
where $\bar{\pi}$ is a set of region constants,
$\bar{\rho}$ is a set of region variables,
$\bar{\varphi}$ is a set of predicate variables,
$regionDeltaMap : \bar{\rho} \rightarrow 2^{\bar{\pi}}$ is a function that
specifies a unification domain for each region variable,
$\predDeltaMap : \bar{\varphi} \rightarrow 2^{\bar{\pi}}$ is a function that
specifies the domain for each predicate variable,
and $C$ is a set of validity constraints, where the $i$'th validity constraint
assumes one of the following forms:
\begin{center}
\(
    \isvalid{\phictxt^{i} \conj \varphi}
            {\phicstr^{i}}\qquad
    \isvalid{\phictxt^{i} \conj \varphi}
            {F_i(\varphi)}
\)
\end{center}
The solution to the constraint satisfaction problem is a pair $(\regionSubstFn,\predSubstFn)$,
where $\regionSubstFn$ is a map from $\rhobar$ to $\bar{\pi}$
and $\predSubstFn$ is a map from $\bar{\varphi}$ to a region-constraint formula such that
\begin{itemize}
  \item $\regionSubstFn(\rho) \in \regionDeltaMap(\rho)$, for every $\rho \in \bar{\rho}$,

  \item $\predSubstFn$ is well-formed under $\predDeltaMap$
    (i.e., $\tywf{\predDeltaMap(\varphi)}{\predSubstFn(\varphi)}$, for every $\varphi \in \bar{\varphi}$).

  \item Every sequent in $C$ is valid after substitutions $\predSubstFn$ and $
    \regionSubstFn$.
  %% $C\lbrack (\regionSubstFn,\predSubstFn) \rbrack$ is valid.
\end{itemize}
\end{definition}

\textbf{TODO: The following not yet integrated into above.}

@@

Solving the constraints is an iterative process, consisting of the following
steps, and this process fails at any point if it determines that no solution
exists for the set of constraints:

\begin{enumerate}

\item
    A constraint of the form $\isvalid{\varphi}{\rho_i = \rho_j}$ unifies
$\rho_i$ and $\rho_j$ if at least one of them is a region variable.
If $\rho_i$ and $\rho_j$ are distinct region constants, then the constraint
solver fails.

\item
When a region variable $\rho$ is unified with a region constant $\pi$,
if $\pi \in \regionDeltaMap(\rho)$, we update $\regionSubstFn(\rho)$ to be $\pi$,
and replace every occurrence of $\rho$ by $\pi$ in the set of constraints.
If $\pi \not\in \regionDeltaMap(\rho)$, the constraint solver fails.

\item
Consider any constraint of the form $\isvalid{\varphi}{\pi_i \outlives \pi_j}$,
where $\pi_i$ and $\pi_j$ are both region constants.
If $\{ \pi_i, \pi_j \} \subseteq \predDeltaMap(\varphi)$, then we simply add
$\pi_i \outlives \pi_j$ as an additional conjunct to $\predSubstFn(\varphi)$.
Otherwise, the constraint solver fails, since no valid solution is possible.

\item
  When any outlives constraint $\phi$ is added to $\predSubstFn(\varphi)$,
  then for every existing validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\varphi)}$,
  we produce a new validity constraint $\isvalid{\varphi' \conj \phictxt}{F(\phi)}$.

%% \emph{Class Constraints}. 
%% Computing the class constraint $\varphi_{\C{C}}$ corresponding to a class \C{C} is
%% fairly straightforward. It is the conjunction of
%% \begin{itemize}
%% \item $\conj_{1 \leq i \leq n} (\rho_i \outlives \rho_0)$
%% \item $\varphi_{\C{B}}$
%% \item $[\rho_a/\rho_f]\varphi_{\C{F}}$
%% \end{itemize}

@@
